Estimation of flow trajectories in a multi-lines network

### 2023-02-09 latex_files/applied_network_science_article/article_v0.tex
rewriting the method (in progress)

%% --------------------------------- OLD IMPORT

@@

 For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$ as
\begin{equation}
	\label{flow_ratio}
	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace ,
\end{equation}
where $r_{ij} > 1$ denotes an overflow through edge $(i, j)$. For a given origin-destination $[s,t]$, define the \emph{orgin-destination flow ratio} $\bar{r}_{st}$ 
as the largest $r_{ij}$ among edge flows $(i,j)\in [s,t]$, that is as 
\begin{equation}
	\label{st_flow_ratio}
	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij}\: \ge\: 1\enspace ,
\end{equation}
where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To adjust the flow, we shall divide the previous flow by this ratio
\begin{equation}
	\label{update_flow}
	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
\end{equation}
and define the new prior distribution  as
\begin{equation}
	\label{update_distrib}
	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}\enspace. 
\end{equation}
where $\phi_s$ and $\psi_t$ are the values (\ref{Sol_LagMult}) obtained in the previous maximum entropy step.

\section{Single line}
\label{Single line}
Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e. $F(i)=i+1$. Define $g_{st}=\frac{1}{(l-1)(l-2)}\chi(s<t)$ (where $\chi(.)$ denotes the 0/1  indicator function) as the maximally uniform prior, reflecting only the unidirectional nature of trips. 
Constraints $n_{r\bullet}=a_r$ (resp. $n_{\bullet r}=b_r$), correspond to $o^r_{st}=
\delta_{sr}$ (resp. $o^r_{st}=\delta_{tr}$) in (\ref{MaxEnt}), which finally yields, after reparametrization, the maximum entropy flow reads
\begin{equation}
\label{nosignle}
n^0_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s<t}c_s\, d_t=n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{equation}
In addition, the (dis-)embarking constraints  yield
\begin{equation}
\label{(dis embarking constraints}
a_i=c_i \sum_{t>i}d_t=c_i\, D_i 
\qquad\qquad\qquad
b_j=d_j\sum_{s<j}c_s=d_j \, C_j  
\end{equation}
where $D_i=\sum_{t>i}d_t$ and $C_j=\sum_{s<j}c_s$. This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow. Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.  Also, 
(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.

@@

\section{Multiple lines}

*** $a_{\bullet}=b_{\bullet}$ (where ``$\bullet$" denotes summation over the replaced index).

@@

%% --------------------------------- FORMALISM

\section{Formalism}

\subsection{Notations}

\subsection{Problem}

\subsection{Solution}

>>>>

%% --------------------------------- NOTATIONS AND FORMALISM

@@

 For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$ as
\begin{equation}
	\label{flow_ratio}
	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace,
\end{equation}
where $r_{ij} > 1$ denotes an overflow through edge $(i, j)$. For a given origin-destination $[s,t]$, define the \emph{orgin-destination flow ratio} $\bar{r}_{st}$ 
as the largest $r_{ij}$ among edge flows $(i,j)\in [s,t]$, that is as 
\begin{equation}
	\label{st_flow_ratio}
	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij}\: \ge\: 1\enspace.
\end{equation}
By construction, $\bar{r}_{st} > 1$ denotes an overflow on some transfer edge between $s$ and $t$. To adjust the flow, we shall divide the previous flow by this ratio
\begin{equation}
	\label{update_flow}
	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
\end{equation}
and define the new prior distribution  as
\begin{equation}
	\label{update_distrib}
	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}\enspace. 
\end{equation}
where $\phi_s$ and $\psi_t$ are the values (\ref{Sol_LagMult}) obtained in the previous maximum entropy step.

\subsection{The case of a single line}
\label{Single line}
A ``network" made of a single line contains no transfers, and flow estimates can be obtained at once by the maximum entropy step only.

Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e. $F(i)=i+1$. The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), reflecting only the unidirectional nature of trips. The margins of the empirical distribution $f_{st}$ as well as the total flow are know, namely 
\begin{displaymath}
\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{displaymath}
and (\ref{Sol}) yields, after reparameterization, maximum entropy flows of the form
\begin{equation}
\label{nosignle}
n_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s}c_s\, \sum_{t>s}d_t=n_{\bullet\bullet}
\end{equation}
where the constraints (\ref{Sol_LagMult}) equivalently express as 
\begin{equation}
\label{dis embarking constraints}
c_s=\frac{a_s}{\sum_{t>s}d_t}
\qquad\qquad\qquad
d_t=\frac{b_t}{\sum_{s<t}c_s}
\end{equation}
to be solved by iterative fitting.

The form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} \cite{wilson1967statistical}  \cite{erlander1990gravity}, where 
$c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to line orientation, but otherwise constant. This constancy entails the following Markovian behavior of flows: let $m_{st}$ be the number of travelers embarking on $s$ and still present at stop $t>s$, and $\rho_s$ the probability to disembark at $s$. Then, by definition and (\ref{nosignle}), ****
\begin{displaymath}
n_{st}=m_{st}\rho_t \qquad\qquad n_{s,t+1}=m_{st}(1-\rho_t)\rho_{t+1}\qquad\qquad \frac{n_{s,t+1}}{n_{st}}=\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=\frac{d_{t+1}}{d_t}
\end{displaymath}
Ainsi 
\begin{displaymath}
\frac{d_{t+2}}{d_{t+1}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}
\qquad\qquad \frac{d_{t+2}}{d_{t}}=\frac{d_{t+2}}{d_{t+1}}\frac{d_{t+1}}{d_{t}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=
\rho_{t+2}(1-\rho_{t+1})(1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
\frac{d_{t+q}}{d_{t}}=
\rho_{t+q}(1-\rho_{t+q-1})(1-\rho_{t+q-2})\ldots(1-\rho_{t+1}) (1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
d_l+d_{l-1}=d_2[\rho_l(1-\rho_{l-1})+\rho_{l-1}](1-\rho_{l-2})(1-\rho_{l-3})\ldots(1-\rho_{3}) (1-\rho_{2})\frac{1}{\rho_{2}}
\end{displaymath}
\begin{displaymath}
D_i=\sum_{t>i}d_t
\end{displaymath}
Soit $z_{st}$ la proportion de voyageurs présents dans le bus en $s$ et toujours présents en $t>s$. Soit $\tau_s$ la probabilité de monter en $s$. Alors
$f_{st}=\tau_sz_{st}\rho_t$

*****

where $D_s=\sum_{t>s}d_t$ and $C_t=\sum_{s<t}c_s$. This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow. Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.  Also, 
(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.

@@

@@
